A Plasma Torus Around a Young Low-Mass Star

### 2025-03-13 paper/ms.tex
update

{\it Age: Lithium}---The depletion of lithium due to $^7$Li(p,
$\alpha$)$^4$He burning in the cores of low-mass stars has been
studied for over sixty years
\cite{Hayashi1963,Bildsten1997,Burke2004}.  \cite{Wood2023} provided
an instructive recent overview: an abbreviated summary is that cooler
and younger M dwarfs can show the 6707.8\,\AA\ doublet in absorption,
$\gtrsim$10\%
below their ``continua''.  Unlike Sun-like stars, the continuum for M
dwarfs is poorly defined due to their molecular absorption.  We
attempted a lithium measurement by constructing a
wavelength-binned and Doppler-corrected TIC~141146667 spectrum,
assigning its uncertainties based on the measured scatter across the
five hour dataset.  We then compared this average spectrum against the
nearest matching template from \cite{Bochanski2007}.  The data show a
slight depression near the
expected lithium wavelength, potentially consistent with a $\Delta
\lambda$$\approx$3\,\AA\ line broadening.  This feature nominally
yields ${\rm EW}_{\rm Li}$=71$^{+18}_{13}$\,m\AA, where the statistical
uncertainties are evaluated using a bootstrap resampling technique
from the statistical uncertainties in the spectrum.  Unfortunately,
the systematic uncertainties associated with the continuum
normalization are large, comparable to the amplitude of the feature;
we therefore prefer to quote the result of this measurement as a
$2\sigma$ upper limit: ${\rm EW}_{\rm Li}$$<$107\,m\AA.
% FIXME: this is perhaps a hokey way to do it.  Better to quantify the
% systematic uncertainty, and incorporate that into your quoted
% value...

Desite uncertainties in the details, what can be stated with
confidence is that lithium is not abundant in the spectrum of
TIC~141146667.  Figure~\ref{fig:liew_population} compares our upper
limit against the lithium equivalent width measurements and
temperatures reported by \cite{Jeffries2023} based on the Gaia-ESO
spectroscopic survey.  If the star were $\lesssim$20\,Myr old, given
its temperature we would expect to see lithium in abundance
($\gtrsim$400\,m\AA).  Since we do not, we conclude that $t_{\rm
Li,emp}$$\gtrsim$20\,Myr.  The \cite{Feiden2016} isochrones provide a
point for theoretical comparison, suggesting that since $M_{\rm
K}$$=$6.67\,mag, $t_{\rm Li,th}$$\gtrsim$35\,Myr for complete
depletion.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf2.pdf}
  \caption[]{todo... }
  \label{fig:liew_population}
\end{figure}

>>>>

{\it Age: Lithium}---The depletion of lithium due to $^7$Li(p,
$\alpha$)$^4$He burning in the cores of low-mass stars has been
studied for over sixty years
\cite{Hayashi1963,Bildsten1997,Burke2004}.  \cite{Wood2023} provided
an instructive recent overview: an abbreviated summary is that cooler
and younger M dwarfs can show the 6707.8\,\AA\ doublet in absorption,
$\gtrsim$10\% below their ``continua''.  Unlike Sun-like stars, the
continuum for M dwarfs is poorly defined due to their molecular
absorption.  We attempted a lithium measurement by constructing a
wavelength-binned and Doppler-corrected TIC~141146667 spectrum,
assigning its uncertainties based on the measured scatter across the
five hour dataset.  We then compared this average spectrum against the
nearest matching template from \cite{Bochanski2007}.  The data show a
small depression near the expected lithium wavelength, potentially
consistent with a $\Delta \lambda$$\approx$3\,\AA\ line broadening.
This feature nominally yields ${\rm EW}_{\rm
Li}$=71$^{+18}_{13}$\,m\AA, where the statistical uncertainties are
evaluated using a bootstrap resampling technique from the statistical
uncertainties in the spectrum.  The systematic uncertainties
associated with the continuum normalization are larger, and likely
comparable to the amplitude of the feature; we therefore treat the
result of this measurement as a $2\sigma$ upper limit: ${\rm EW}_{\rm
Li}$$<$107\,m\AA.

Desite uncertainties in the details, what can be stated with
confidence is that lithium is not abundant in the spectrum of
TIC~141146667.  Figure~\ref{fig:liew_population} compares our upper
limit against the lithium equivalent width measurements reported by
\cite{Jeffries2023} based on the Gaia-ESO spectroscopic survey.  If
the star were $\lesssim$20\,Myr old, given its temperature we would
expect to see lithium in abundance ($>$400\,m\AA).  Since we do not,
we conclude that $t_{\rm Li,emp}$$\gtrsim$20\,Myr.  The
\cite{Feiden2016} isochrones provide a point for theoretical
comparison, suggesting that since $M_{\rm K}$$=$6.67\,mag, $t_{\rm
Li,th}$$\gtrsim$35\,Myr would be expected for complete depletion (see
e.g.~Figure 7 from \cite{Wood2023}).

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf2.pdf}
  \caption{An upper limit on photospheric lithium for TIC~141146667
  (green triangle) yields a lower limit on the star's age of $\gtrsim$20\,Myr.
  Comparison stars are from the Gaia-ESO survey
  \cite{Jeffries2023}; rich clusters in each age bin include
  NGC\,2264 (4.5\,Myr), $\lambda$~Ori (8.7\,Myr), ${\rm \gamma}^2$\,Vel (16.4\,Myr), NGC\,2547
  (35.0\,Myr), IC\,2602 and IC\,2391 (42.0\,Myr), and NGC\,2451A
  (50.0\,Myr). }
  \label{fig:liew_population}
\end{figure}
